Photo-D with LSST: stellar photometric distances out to the edge of the Galaxy

### 2024-01-19 paper/analysis.tex
more work on Methods

>>>>

XXX moved from Methods, use when discussing 2-D posteriors...

Another important degeneracy arises from the fact that even for a fixed \teff and \feh, the \logg and thus the luminosity are not uniquely determined by the colors: a degeneracy may exist between the giant branch and the main sequence as the colors constructed from \textit{ugrizy} bands are not sensitive to \logg.

### 2024-01-19 paper/discussion.tex
more work on Methods

>>>>

XXX  old text from methods 
The best photometric estimators of metallicity are colors whose shorter-wavelength component includes the metal absorbtion bands at near-UV wavelengths, short of Balmer break (300 $\lessapprox\lambda$ [nm] $\lessapprox$ 400). Therefore, the LSST has a comparative advantage over the surveys lacking \textit{u}-band measurements, and could provide accurate distances within the range of 5-10\%. \magcom{A plot of model spectra, fixed, \textit{log(g)} and \textit{T\textsubscript{eff}}, several different metallicities?}

Likelihood computations utilize MIST/Dartmouth isochrones
and priors are derived from TRILEGAL-based simulated LSST catalogs from Dal Tio et al. (2022). The computation speed is about 10 milliseconds per
star on a single core for both optimized grid search and MCMC methods; we show in a companion paper by Mrakov\v{c}i\'{c} et al. how to utilize neural
networks to accelerate this performance by an order of magnitude.  We validate our pipeline, named
PhotoD (in analogy with photo-z, photometric redshifts of galaxies) using both simulated catalogs, and SDSS, DECam and Gaia data. 
We intend to make LSST-based value-added PhotoD catalogs publicly
available via Rubin Science Platform with every LSST Data Release.

In this paper we describe a method that will deliver LSST-based stellar distance estimations complementary to \textit{Gaia's} state-of-the-art trigonometric parallaxes and reach about 10 times further, to approximately 100 kpc. These results will be transformative for the studies of the Milky Way in general, and the stellar and the dark matter halo in particular as never before was there a survey that simultaneously observed roughly two thirds of the sky, to the co-added depth of \textit{r}$\approx$26 mag.

\magcom{A bit about the importance of the distance estimation in the MW, dust implications (for extragalactic science too).}

There are a variety of astronomical methods to estimate distances to stars, ranging from direct geometric (trigonometric) methods for nearby stars to indirect methods based on astrophysics for more distant stars.

\magcom{Mention \cite{bailer-jones_estimating_2021}, \cite{gordon_panchromatic_2016}, \cite{green_measuring_2014,green_three-dimensional_2015,green_3d_2019}, \cite{juric_milky_2008} and \cite{lallement_3d_2014}, \cite{queiroz_starhorse_2018}}.

Layout of the paper is...

 assuming approximately 10 billion stars in the LSST stellar catalog as predicted by TRILEGAL, and a 100-core machine, a distance catalog could be produced in about 10 days (10 millisec/star). Input LSST photometry (with errors) and minimalistic auxiliary metadata  for 10 billion stars would amount to about 1 TB of data; the code outputs, including a covariance matrix of the three fitted parameters, would have approximately the same volume.

 Thanks to the \textit{Vera C. Rubin} observatory's \textit{Legacy Survey of Space and Time} (LSST), for the first time in history, an astronomical catalog will contain more Milky Way stars than there are living people on Earth -- of the order 10-20 billion, depending on model assumptions. In order to map the Milky Way in three dimensions, distances to these stars must be accurately  estimated.

### 2024-01-19 paper/intro.tex
Minor edits to the text.
more work on Methods

Thanks to the \textit{Vera C. Rubin} observatory's \textit{Legacy Survey of Space and Time} (LSST), for the first time in history, an astronomical catalog will contain more Milky Way stars than there are living people on Earth -- of the order 10-20 billion, depending on model assumptions. In order to map the Milky Way in three dimensions, distances to these stars must be accurately  estimated. In this paper we describe a method that will deliver LSST-based stellar distance estimations complementary to \textit{Gaia's} state-of-the-art trigonometric parallaxes and reach about 10 times further, to approximately 100 kpc. These results will be transformative for the studies of the Milky Way in general, and the stellar and the dark matter halo in particular as never before was there a survey that simultaneously observed roughly two thirds of the sky, to the co-added depth of \textit{r}$\approx$26 mag.

\magcom{A bit about the importance of the distance estimation in the MW, dust implications (for extragalactic science too).}

There are a variety of astronomical methods to estimate distances to stars, ranging from direct geometric (trigonometric) methods for nearby stars to indirect methods based on astrophysics for more distant stars.

\magcom{Mention \cite{bailer-jones_estimating_2021}, \cite{gordon_panchromatic_2016}, \cite{green_measuring_2014,green_three-dimensional_2015,green_3d_2019}, \cite{juric_milky_2008} and \cite{lallement_3d_2014}, \cite{queiroz_starhorse_2018}}.

Layout of the paper is...

>>>>

In order to map the Milky Way in three dimensions, distances to its stars must be accurately  estimated.
Kinematic studies based on proper motion data also require estimates of stellar distances.  There are a variety
of astronomical methods to estimate distances to stars, ranging from direct geometric (trigonometric) methods
for nearby stars to indirect methods based on astrophysics for more distant stars.

As demonstrated with SDSS \citep{2008ApJ...673..864J, 2008ApJ...684..287I}, Pan-STARRS \citep{2014ApJ...783..114G, green_3d_2019} and most recently with
Gaia data \citep{bailer-jones_estimating_2021}, broad-band near-UV to near-IR stellar photometry can be used 
to estimate distance, metallicity and interstellar dust extinction along the line of sight for stars in the Galaxy.
In analogy with photo-z, photometric redshifts of galaxies, hereafter we refer to these methods as photo-D. 
The photo-D method is conceptually quite simple: multi-dimensional color tracks (either empirical or model-based),
parametrized by luminosity, metallicity and extinction, are fit to observed colors and the best-fit produces estimates
of the three model parameters. The method relies on strong correlations between stellar colors and
stellar luminosity for dominant stellar populations such as main-sequence stars, red giants, white dwarfs, and even
for the majority of unresolved binary stars. These are the same correlations that are responsible for the abundant
structure seen in the Hertzsprung-Russell diagram. In addition to colors and luminosity, these correlations
also involve metallicity in case of main-sequence and red giant stars, and surface gravity in case of white dwarfs.
For the youngest main-sequence stars, stellar age may play a role, too. Measured stellar colors are also affected
by interstellar dust extinction along the line of sight towards the star. Consequently, measurements of apparent
brightness and sufficient number of UV to IR colors, such as those that Rubin Observatory's Legacy Survey of Space
and Time (LSST; \citealt{2019ApJ...873..111I}) will provide, can be used to accurately estimate these parameters,
and ultimately stellar distances.

LSST-based stellar distance estimates will significantly improve available distance catalogs, such as those recently produced by \cite{green_3d_2019}
and \cite{bailer-jones_estimating_2021}. First, the sample size will be increased by more than an order of magnitude, and exceed 10 billion stars.
Distance accuracy for stars with sufficiently small photometric errors will be within the 5-10\% range, or about twice
as accurate as for surveys lacking the UV $u$ band (which provides metallicity constraints). LSST-based stellar distances
will reach about 10 times further than Gaia’s color-based distances and will be transformative for studies of the
Milky Way in general, and of its stellar and dark matter halo in particular.

In this paper we present a Bayesian procedure and pipeline that build on previous work and can handle LSST-sized datasets.
In \S2, we describe methodology and in \S3 we test the pipeline using both simulated catalogs, and SDSS, DECam and Gaia
photometry. We discuss possibilities for further improvements and catalog public release plans in \S4.

### 2024-01-19 paper/paper.tex
more work on Methods

\title{PhotoD with LSST: stellar photometric distances out to the edge of the Galaxy}

@@

\author[0000-0000-0000-0000]{Bob Abel}
\affiliation{The Paradise Island, WA, USA }

@@

\author[0000-0000-0000-0000]{Filip Matkovi\'{c}}
\affiliation{Hvar Observatory, 10000 Zagreb, Croatia}

@@

\author[0000-0000-0000-0000]{Toni \v{S}ari\'{c}}
\affiliation{University of Split, Croatia}

\author[0000-0000-0000-0000]{Tomislav Jurki\'{c}}
\affiliation{University of Rijeka, Croatia}

@@

\author[0000-0000-0000-0000]{Forgotten Collaborator}
\affiliation{Somewhere}

@@

%% Mark off the abstract in the ``abstract'' environment. 
\begin{abstract}
As demonstrated with SDSS, Pan-STARRS and most recently with Gaia data, broad-band near-UV to near-IR stellar photometry can be used 
to estimate distance, metallicity and interstellar dust extinction along the line of sight for stars in the Galaxy. Anticipating
photometric catalogs with tens of billions of stars from Rubin's Legacy Survey of Space and Time (LSST), we present a Bayesian procedure 
and pipeline that build on previous work and can handle LSST-sized datasets. Likelihood computations utilize MIST/Dartmouth isochrones
and priors are derived from TRILEGAL-based simulated LSST catalogs from Dal Tio et al. (2022). The computation speed is about 10 milliseconds per
star on a single core for both optimized grid search and MCMC methods; we show in a companion paper by Mrakov\v{c}i\'{c} et al. how to utilize neural
networks to accelerate this performance by an order of magnitude.  We validate our pipeline, named
PhotoD (in analogy with photo-z, photometric redshifts of galaxies) using both simulated catalogs, and SDSS, DECam and Gaia data. 
We intend to make LSST-based value-added PhotoD catalogs publicly
available via Rubin Science Platform with every LSST Data Release. 
\end{abstract}

@@

%\include{methodologyZI}

>>>>

\title{Photo-D with LSST: stellar photometric distances out to the edge of the Galaxy}

@@

\author[0000-0001-8418-3083]{Bob Abel}
\affiliation{Olympic College, 1600 Chester Avenue, Bremerton, WA
98337, USA, and DiRAC Institute, University of Washington, Box 351580, Seattle, WA 98195, USA}

@@

\author[0009-0002-5858-585X]{Filip Matkovi\'{c}}
\affiliation{Hvar Observatory, Faculty of Geodesy, University of Zagreb, Ka\v{c}i\'{c}eva 26, 10000 Zagreb, Croatia}

@@

\author[0000-0001-8731-8369]{Toni \v{S}ari\'{c}}
\affiliation{University of Split - FESB, R. Bo\v{s}kovi\'{c}a 32, 21000 Split, Croatia}

\author[0000-0000-0000-0000]{Tomislav Jurki\'{c}}
\affiliation{Faculty of Physics, University of Rijeka, Radmile Matej\v{c}i\'{c} 2, 51000 Rijeka, Croatia}

@@

%\author[0000-0000-0000-0000]{Forgotten Collaborator}
%\affiliation{Somewhere}

@@

%% Mark off the abstract in the ``abstract'' environment. 
\begin{abstract}
As demonstrated with SDSS, Pan-STARRS and most recently with Gaia data, broad-band near-UV to near-IR stellar photometry can be used 
to estimate distance, metallicity and interstellar dust extinction along the line of sight for stars in the Galaxy. Anticipating
photometric catalogs with tens of billions of stars from Rubin's Legacy Survey of Space and Time (LSST), we present a Bayesian procedure 
and pipeline that build on previous work and can handle LSST-sized datasets. Likelihood computations utilize MIST/Dartmouth isochrones
and priors are derived from TRILEGAL-based simulated LSST catalogs from Dal Tio et al. (2022). The computation speed is about 10 milliseconds per
star on a single core for both optimized grid search and MCMC methods; we show in a companion paper by Mrakov\v{c}i\'{c} et al. how to utilize neural
networks to accelerate this performance by an order of magnitude.  We validate our pipeline, named
PhotoD (in analogy with photo-z, photometric redshifts of galaxies)
using both simulated catalogs, and SDSS, DECam and Gaia photometry. We intend to make LSST-based value-added PhotoD catalogs publicly
available via Rubin Science Platform with every LSST Data Release. 
\end{abstract}

@@
